Depth refugia for reef persistence in the GBR

### 2025-10-20 paper/paper.qmd
Incorporate Sam's final changes

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

@@

engine: julia
julia:
  exeflags: ["+1.11.5"]

@@

editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

The GBR is the world's largest coral reef ecosystem consisting of thousands of individual
reefs interconnected through larval dispersal pathways mediated by hydrodynamic, spatial,
and ecological processes [@cappo2000]. The GBR faces severe projected coral cover declines
under climate change [@bozec2025; @bozec2022; @condie2021]. Coral cover declines will be
compounded by ocean acidification, tropical cyclones, and crown-of-thorns starfish outbreaks
[@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24
Summer mass bleaching event may be signalling a shift towards this reef future, with some
level of bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs),
representing the most widespread bleaching event on the GBR and largest single year decline
in the history of the Long Term Monitoring Program (LTMP;
@australianinstituteofmarinescience2024; @cantin2024).

@@

The spatial extent of reefs represented in this investigation follows reef definitions
available via the Great Barrier Reef Marine Park Authority (GBRMPA). Reef definitions
outline 3806 reefs across the GBR [@GBRMPA2022] (@fig-region-map). Prior analyses by Bozec
et al. [-@bozec2022] provided estimates of coral carrying capacity for each reef that were
adopted by this study.

@@

These GCMs were selected based on evaluation by Grose et al., [-@grose2023], indicating
their ability to recreate historical atmospheric temperature and precipitation conditions in
the Australasian region. They span a range of Equilibrium Climate Sensitivity (ECS) values
within the Intergovernmental Panel on Climate Change (IPCC) assessed likely range. The IPCC
defines ECS as the long-term warming caused by a doubling of carbon dioxide above its
pre-industrial concentration [@forster2021]. While Grose et al. [-@grose2023] do not
consider oceanic data, they provide a region-specific selection mechanism which is used as
an indicator of general performance of GCMs for the region. An overview of the selected GCMs
is provided in @tbl-gcms. Note here that Grose et al., [-@grose2023] assessed GFDL-ESM4,
whereas Dixon et al., [-@dixon2022; -@dixon2023] provide data based on GFDL-CM4. This
substitution is assumed appropriate given they share the same fundamental representation of
the physical climate system. The use of these GCMs therefore allows for a range of possible
futures to be explored, capturing uncertainty in future responses to CO~2~ increases
(@fig-methods-dhws; @sfig-ecs; @grose2023; @shepherd2018) and enables the assessment of
modelled coral reef behaviour over a broad range of independently derived climate futures
[@abramowitz2019; @brunner2020].

@@

Information for each GCM used in this investigation including Equilibrium Climate
Sensitivity (ECS) value, interpreted relative severity and where results for an individual
GCM results are found. GCMs in context with assessed plausible ranges of ECS values are
displayed in [@sfig-ecs]. ACCESS-ESM1-5 has an ECS closest to the midpoint of the likely ECS
range and was taken as an intermediate model with results displayed in the main text.
:::

@@

The coral larval connectivity model used in this study comprises 3D hourly velocities from
the 1 km grid resolution hydrodynamic model (GBR-1 version 2.0; @steven2019), as input to
OceanParcels [@delandmeter2019], a particle tracking model. The connectivity model was used
to simulate larval dispersal and settlement of broadcast-spawning Acropora corals in the
GBR, following annual mass spawning events in 2015, 2016, 2018, 2019 and split-spawning
events in 2017 and 2020. Split spawning occurs when the annual reproductive event of a coral
population is split over two consecutive months. Coral mass spawning usually occurs 4-6 days
after the full moon during the spring-summer transition in the GBR with variations across
species and locations. The spawning days identified are: 2015 - November 30, December 1 and
2; 2016 - November 18, 19 and 20; 2017 - November 8, 9 and 10, December 8, 9 and 10; 2018 –
November 18, 19, 20; 2019 – November 16, 17, 18; 2020 – November 5, 6 and 7, December 4, 5,
6.

For each spawning night, particles were released from random locations within each reef
polygon as neutrally buoyant virtual larvae every 27 seconds from 8:00 pm to 11:00 pm at
2.25 m below the surface. In each of the reef polygons, 401 particles were released
resulting in a total of 1,526,206 particles every night of spawning across the GBR. The
model was run for 30 days with a 3-minute timestep, and the locations of the released
particles were updated every 15 minutes until the end of the simulation. Larvae were assumed
to become competent to settle from 4 to 28 days post release with a daily mortality rate of
0.4 [@ani2024], which was implemented in 15-minute timestep as probabilistic mortality,
corresponding to 0.0053 mortality rate per timestep. To incorporate the stochastic sensory
behaviour of larvae during settlement, a 0.01 probability of settlement is assumed when
larvae are over a reef during the competency period per 15-minute timestep. The information
obtained on larvae settlement was used to compute a connectivity matrix among reef polygons
for each spawning night (see Ani et al. [-@ani2024] for more details). A total of 24
connectivity matrices were generated for the spawning nights that were considered.

@@

Time series clustering was applied over the time frame between 2025 and 2075. The clustering
was conducted at three spatial scales: (1) the whole of GBR; (2) each GBR management area;
and (3) for each reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were
conducted at the different spatial scales as, conceptually, the temporal behaviour
constituting "low", "medium" or "high" coral cover may differ at different spatial scales.
Analyses at the whole-of-GBR scale were found to lead to identical conclusions regarding
reef characteristics which influence coral cover and so for brevity are not shown here.

@@

There were substantial differences in the maintenance of coral cover post 2030 between the
GCMs (@fig-gcm-cover-trajectories). Of the GCMs explored, NorESM2-MM had the lowest ECS
(2.5°C) and reefs exhibit continuing recovery and decline dynamics up to 2075 under this
hypothetical climate future. Unsurprisingly, recovery phases were predominantly seen in the
reefs in the High cover cluster, with Medium and Low exhibiting lesser probability of such
periods of recovery. The High cover cluster exhibits greater resilience despite encompassing
a broad range of initial coral cover conditions at 2025 (\@fig-gcm-cover-trajectories).
Generally, recovery and increases in coral cover were visible in simulated results up to
2045. Under NorESM2-MM, the Far Northern Management area was found to have the most similar
temporal behaviour between Low and High clusters compared to other Management areas
(@stbl-cluster-cid) due to the low degree of coral cover overall
(@fig-gcm-cover-trajectories).

@@

Reef median depth had a moderate positive correlation with the number of years a reef
maintains coral cover above the assumed carbonate budget threshold required to sustain live
coral cover ($\theta_{0}$). Greater median depth was a key characteristic defining reefs
that sustain coral cover above $\theta_{0}$ (@fig-access-esm1-5-depth-conn-carbonate;
@sfig-access-esm1-5-carb-budget-corr; @sfig-access-esm1-5-mean-carb-budget). Unsurprisingly,
the number of years above $\theta_0$ substantially decreased as $\theta_0$ is increased, as
does the Spearman's correlation (@sfig-access-esm1-5-carb-budget-corr). Decreasing
correlation levels are likely due to an increasing number of reefs being unable to maintain
coral cover above $\theta_0$ as it is increased, with the distribution of years above
threshold becoming increasingly right-skewed (@fig-access-esm1-5-depth-conn-carbonate). In
contrast, weighted incoming connectivity strength (log10 scale) was found to have no
correlation with the number of years a reef maintains cover above the required thresholds
(@sfig-access-esm1-5-carb-budget-corr).

@@

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. No direct, first order, relationship between weighted incoming connectivity
and the ability of reefs to maintain positive carbonate budgets under future conditions was
found. There has been contrasting evidence [@bode2018; @hock2017] as to whether current
connectivity modelling is able to identify reefs that are important for recovery processes,
with findings dependent on the larval connectivity model used. These investigations were
performed using prior generations of larval connectivity models parameterised for 2008-2013
[@hock2017] and 1996-2002 [@bode2018] time periods, without considering the influence of
reef depth on resilience. Using updated 1km^2^ resolution larval connectivity across the
whole GBR, with a limited temporal sample of particle flows, it was found that larval
connectivity alone may not be important or reef resilience and recovery but should be
considered in combination with more influential factors such as depth.

Findings on the importance of connectivity likely differ from literature due to the highly
uncertain nature of larval connectivity modelling [@choukroun2025; @vogt-vincent2023] and
the way larval flows and juvenile recruitment are implemented in ADRIA - CoralBlox.
Additionally, and the high magnitude of early bleaching events (2025-2035) may decrease the
breeding capacity of strongly connected reefs, resulting in reduced importance of
connectivity in recovery. Analysing daily connectivity data Vogt-Vincent, Mitarai and
Johnson [-@vogt-vincent2023] suggest that time-mean connectivity may not be suitable for
capturing the seasonal - ecological time scale connectivity between locations due to high
daily variability in connectivity data. Connectivity matrices used in this investigation
have large relative standard deviations for larval transition probabilities between reefs
over the 24 days used (@sfig-connectivity-rsd). Using alternative summarisation methods that
capture ecological-timescale patterns in coral larval dispersal, such as cumulated implicit
connectivity [@vogt-vincent2023], future investigations may find increased importance of
both depth and reef connectivity for resilience, in line with other investigations
[@hock2017; @torres2024]. Choukroun et al [-@choukroun2025] performed a comparison of three
GBR-scale connectivity models, including GBR1, and suggest the use of a model ensemble
approach to estimate larval dispersal due to the high sensitivity of connectivity
predictions to model specification. Further improvements and research in connectivity
modelling are necessary when incorporating connectivity as a supplementary factor in reef
resilience and intervention-planning analyses.

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. Coral colonies are modelled as diameter values that increase over time
through growth. Additionally, the influence of light intensity on coral growth or the
negative influence of depth on coral growth rates due to reduced available light levels are
not modelled in this investigation [@baker1975; @huston1985; @ruiz-diaz2022]. If included in
this investigation, it is possible that reef recovery ability, via regrowth, would be
reduced at greater depths, reducing the positive effect of depth refugia for temporal
resilience. Additionally, reduced growth rates at increased depths may reduce carbonate
budget maintenance due to lower rates of CaCO~3~ accretion. Due to the availability of
quality future projected data this investigation does not consider other important impacts
such as waves, cyclones, ocean currents and Crown of Thorns Starfish (CoTS) on coral colony
survival and bleaching [@massel1993; @deaker2022].

@@

In addition to the usual challenges of deriving data from satellite imagery, it is
acknowledged that there is high potential for misalignment between the bathymetry data and
GBRMPA reef outlines. Hammerton and Lawrey [-@hammerton2025], for example, identified 120
false positives, where features identified as reefs were sand banks or rocky reefs (rather
than coral reefs). Reefs with bathymetry data covering less than 5% of their indicative area
were excluded from analyses (see @sfig-depth-cutoff). Existing GBRMPA reef outlines were
adopted to align with other ongoing modelling and analyses conducted as part of the Reef
Restoration and Adaptation Program (<https://gbrrestoration.org>; RRAP [-@thereef]).

@@

The full reproducible investigation, and links to required data sources, can be found at
Grier and Iwanaga ([-@grier2025]; https://doi.org/10.5281/zenodo.17333564).

>>>>

---
title: Depth refugia for reef persistence in the GBR

@@

engine: julia
exeflags: ["--project=."]

@@

editor:
  markdown:
    wrap: 92
    canonical: true
citeproc: true
---

@@

The GBR is the world's largest coral reef ecosystem consisting of thousands of individual
reefs interconnected through larval dispersal pathways mediated by hydrodynamic, spatial,
and ecological processes [@cappo2000]. The GBR faces severe projected coral cover declines
under climate change [@bozec2025; @bozec2022; @condie2021]. Coral cover declines will be
compounded by ocean acidification, tropical cyclones, and crown-of-thorns starfish outbreaks
[@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24
Summer mass bleaching event may be signalling a shift towards this reef future, with some
level of bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs),
representing the most widespread bleaching event on the GBR and largest single year decline
in the history of the Long Term Monitoring Program [LTMP,
@australianinstituteofmarinescience2024; @cantin2024].

@@

The spatial extent of reefs represented in this investigation follows reef definitions
available via the Great Barrier Reef Marine Park Authority (GBRMPA). Reef definitions
outline 3806 reefs across the GBR (@fig-region-map; GBRMPA [-@GBRMPA2022]). Prior analyses
by Bozec et al. [-@bozec2022] provided estimates of coral carrying capacity for each reef
that were adopted by this study.

@@

These GCMs were selected based on evaluation by Grose et al., [-@grose2023], indicating
their ability to recreate historical atmospheric temperature and precipitation conditions in
the Australasian region. They span a range of Equilibrium Climate Sensitivity (ECS) values
within the Intergovernmental Panel on Climate Change (IPCC) assessed likely range. The IPCC
defines ECS as the long-term warming caused by a doubling of carbon dioxide above its
pre-industrial concentration [@forster2021]. While Grose et al. [-@grose2023] do not
consider oceanic data, they provide a region-specific selection mechanism which is used as
an indicator of general performance of GCMs for the region. An overview of the selected GCMs
is provided in @tbl-gcms. Note here that Grose et al., [-@grose2023] assessed GFDL-ESM4,
whereas Dixon et al., [-@dixon2022; -@dixon2023] provide data based on GFDL-CM4. This
substitution is assumed appropriate given they share the same fundamental representation of
the physical climate system. The use of these GCMs therefore allows for a range of possible
futures to be explored, capturing uncertainty in future responses to CO~2~ increases
(@fig-methods-dhws; @sfig-ecs; [@grose2023; @shepherd2018]) and enables the assessment of
modelled coral reef behaviour over a broad range of independently derived climate futures
[@abramowitz2019; @brunner2020].

@@

Information for each GCM used in this investigation including Equilibrium Climate
Sensitivity (ECS) value, interpreted relative severity and where results for an individual
GCM results are found. GCMs in context with assessed plausible ranges of ECS values are
displayed in (@sfig-ecs). ACCESS-ESM1-5 has an ECS closest to the midpoint of the likely ECS
range and was taken as an intermediate model with results displayed in the main text.
:::

@@

The coral larval connectivity model used in this study comprises 3D hourly velocities from
the 1 km grid resolution hydrodynamic model [GBR-1 version 2.0, @steven2019], as input to
OceanParcels [@delandmeter2019], a particle tracking model. The connectivity model was used
to simulate larval dispersal and settlement of broadcast-spawning Acropora corals in the
GBR, following annual mass spawning events in 2015, 2016, 2018, 2019 and split-spawning
events in 2017 and 2020. Split spawning occurs when the annual reproductive event of a coral
population is split over two consecutive months. Coral mass spawning usually occurs 4-6 days
after the full moon during the spring-summer transition in the GBR with variations across
species and locations. The spawning days identified are: 2015 - November 30, December 1 and
2; 2016 - November 18, 19 and 20; 2017 - November 8, 9 and 10, December 8, 9 and 10; 2018 –
November 18, 19, 20; 2019 – November 16, 17, 18; 2020 – November 5, 6 and 7, December 4, 5,
6.

For each spawning night, particles were released from random locations within each reef
polygon as neutrally buoyant virtual larvae every 27 seconds from 8:00 pm to 11:00 pm at
2.25 m below the surface. In each of the reef polygons, 401 particles were released
resulting in a total of 1,526,206 particles every night of spawning across the GBR. The
model was run for 30 days with a 3-minute timestep, and the locations of the released
particles were updated every 15 minutes until the end of the simulation. Larvae were assumed
to become competent to settle from 4 to 28 days post release with a daily mortality rate of
0.4 [@ani2024], which was implemented in 15-minute timestep as probabilistic mortality,
corresponding to 0.0053 mortality rate per timestep. To incorporate the stochastic sensory
behaviour of larvae during settlement, a 0.01 probability of settlement is assumed when
larvae are over a reef during the competency period per 15-minute timestep. The information
obtained on larvae settlement was used to compute a connectivity matrix among reef polygons
for each spawning night (see @ani2024 for more details). A total of 24 connectivity matrices
were generated for the spawning nights that were considered.

@@

Time series clustering was applied over the time frame between 2025 and 2075. The clustering
was conducted at three spatial scales: (1) the whole of GBR; (2) each GBR management area;
and (3) for each reefal bioregion (defined in @GBRMPA2024). Analyses were conducted at the
different spatial scales as, conceptually, the temporal behaviour constituting "low",
"medium" or "high" coral cover may differ at different spatial scales. Analyses at the
whole-of-GBR scale were found to lead to identical conclusions regarding reef
characteristics which influence coral cover and so for brevity are not shown here.

@@

There were substantial differences in the maintenance of coral cover post 2030 between the
GCMs (@fig-gcm-cover-trajectories). Of the GCMs explored, NorESM2-MM had the lowest ECS
(2.5°C) and reefs exhibit continuing recovery and decline dynamics up to 2075 under this
hypothetical climate future. Unsurprisingly, recovery phases were predominantly seen in the
reefs in the High cover cluster, with Medium and Low exhibiting lesser probability of such
periods of recovery. The High cover cluster exhibits greater resilience despite encompassing
a broad range of initial coral cover conditions at 2025 (@fig-gcm-cover-trajectories).
Generally, recovery and increases in coral cover were visible in simulated results up to
2045. Under NorESM2-MM, the Far Northern Management area was found to have the most similar
temporal behaviour between Low and High clusters compared to other Management areas
(@stbl-cluster-cid) due to the low degree of coral cover overall
(@fig-gcm-cover-trajectories).

@@

Reef median depth had a moderate positive correlation with the number of years a reef
maintains coral cover above the assumed threshold required to sustain a positive carbonate
budget ($\theta_{0}$). Greater median depth was a key characteristic defining reefs that
sustain coral cover above $\theta_{0}$ (@fig-access-esm1-5-depth-conn-carbonate;
@sfig-access-esm1-5-carb-budget-corr; @sfig-access-esm1-5-mean-carb-budget). Unsurprisingly,
the number of years above $\theta_0$ substantially decreased as $\theta_0$ is increased, as
does the Spearman's correlation (@sfig-access-esm1-5-carb-budget-corr). Decreasing
correlation levels are likely due to an increasing number of reefs being unable to maintain
coral cover above $\theta_0$ as it is increased, with the distribution of years above
threshold becoming increasingly right-skewed (@fig-access-esm1-5-depth-conn-carbonate). In
contrast, weighted incoming connectivity strength (log10 scale) was found to have no
correlation with the number of years a reef maintains cover above the required thresholds
(@sfig-access-esm1-5-carb-budget-corr).

@@

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. No direct, first order, relationship between weighted incoming connectivity
and the ability of reefs to maintain positive carbonate budgets under future conditions was
found. There has been contrasting evidence [@bode2018; @hock2017] as to whether current
connectivity modelling is able to identify reefs that are important for recovery processes,
with findings dependent on the larval connectivity model used. These investigations were
performed using prior generations of larval connectivity models parameterised for 2008-2013
[@hock2017] and 1996-2002 [@bode2018] time periods, without considering the influence of
reef depth on resilience. Using updated 1km^2^ resolution larval connectivity across the
whole GBR, with a limited temporal sample of particle flows, it was found that larval
connectivity alone may not be enough to support reef resilience and recovery but should be
considered in combination with more influential factors such as depth.

Findings on the importance of connectivity likely differ from literature due to the highly
uncertain nature of larval connectivity modelling [@choukroun2025; @vogt-vincent2023] and
the way larval flows and juvenile recruitment are implemented in ADRIA - CoralBlox.
Additionally, and the high magnitude of early bleaching events (2025-2035) may increase the
influence of temperature mediators (i.e. depth) in our simulations decrease the breeding
capacity of strongly connected reefs, resulting in reduced importance of connectivity in
recovery. Analysing daily connectivity data Vogt-Vincent, Mitarai and Johnson
[-@vogt-vincent2023] suggest that time-mean connectivity may not be suitable for capturing
the seasonal - ecological time scale connectivity between locations due to high daily
variability in connectivity data. Connectivity matrices used in this investigation have
large relative standard deviations for larval transition probabilities between reefs over
the 24 days used (@sfig-connectivity-rsd). Using alternative summarisation methods that
capture ecological-timescale patterns in coral larval dispersal in future investigations,
such as cumulated implicit connectivity [@vogt-vincent2023], may find increased importance
of both depth and reef connectivity for resilience, in line with other studies [@hock2017;
@torres2024]. Choukroun et al [-@choukroun2025] performed a comparison of three GBR-scale
connectivity models, including GBR1, and suggest the use of a model ensemble approach to
estimate larval dispersal due to the high sensitivity of connectivity predictions to model
specification. Further improvements and research in connectivity modelling are necessary
when incorporating connectivity as a supplementary factor in reef resilience and
intervention-planning analyses.

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. Coral colonies are modelled as diameter values that increase over time
through growth. Additionally, the influence of light intensity on coral growth or the
negative influence of depth on coral growth rates due to reduced available light levels are
not modelled in this investigation [@baker1975; @huston1985; @ruiz-diaz2022]. If included in
this investigation, it is possible that reef recovery ability, via regrowth, would be
reduced at greater depths, reducing the positive effect of depth refugia for temporal
resilience. Additionally, reduced growth rates at increased depths may reduce carbonate
budget maintenance due to lower rates of CaCO~3~ accretion. Due to the availability of
quality future projected data this investigation does not consider other important impacts
such as waves, cyclones, ocean currents and Crown of Thorns Starfish (CoTS) on coral colony
survival and bleaching [@emslie2024a; @emslie2024b].

@@

In addition to the usual challenges of deriving data from satellite imagery, it is
acknowledged that there is high potential for misalignment between the bathymetry data and
GBRMPA reef outlines. Hammerton and Lawrey [-@hammerton2025], for example, identified 120
false positives, where features identified as reefs were sand banks or rocky reefs (rather
than coral reefs). Reefs with bathymetry data covering less than 5% of their indicative area
were excluded from analyses (see @sfig-depth-cutoff). Existing GBRMPA reef outlines were
adopted to align with other ongoing modelling and analyses conducted as part of the Reef
Restoration and Adaptation Program [<https://gbrrestoration.org>, @thereef].

@@

The full reproducible investigation, and links to required data sources, can be found at
Grier and Iwanaga [-@grier2025, https://doi.org/10.5281/zenodo.17333564].
